---
title: "Reproducible Research Week(4) - Project(2)"
author: "Vijayeshwari Palakurthi"
date: "26 June 2020"
output:
     html_document: default
     pdf_document: default
     word_document: default
---
### **1.OVERVIEW**

  *  Weather events cause public health and economic problems for communities and municipalities. Severe events result in fatalities, injuries, and damage. Predicting and/or preventing these outcomes is a primary objective.

  *  This analysis examines the damaging effects of severe weather conditions (e.g. hurricanes, tornadoes, thunderstorms, floods, etc.) on human populations and the econonomy in the U.S. from 1950 to 2011.

  *  As a result, the analysis will highlight the severe weather events associated with the greatest impact on the economy and population health.
  
### **2.SYNOPSIS**

  *  This is an exploration of the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database.
  *  This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, which type of event, as well as the estimates of relevant fatalities, injuries, and various forms of damage.
  *  The dataset used in this project is provided by the U.S. National Oceanic and Atmospheric Administration (NOAA).
  *  This analysis discovered that tornados are responsible for a maximum number of fatalities and injuries.
  *  This analysis also discoered that floods are responsbile for maximum property damage, while Droughts cause maximum crop damage.

Objective: Explore the NOAA Storm Database to help answer important questions about severe weather events.

### DATA PROCESSING
### DATA PREP
### DP1.1 Install packages & Load libraries

```{r}
library(R.utils)
library(rmarkdown)
library(knitr)
Sys.setlocale("LC_TIME", "English")
```
DP1.2 Download the storm data file into the designated working directory foldeR

```{r}
temp <- tempfile()

##Performing the download
if(!file.exists("/stormData.csv.bz2")){
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile="./stormData.csv.bz2")
}
##Uncompressing the file
if(!file.exists("stormdata.csv"))
{
  unzip("stormData.csv.bz2","stormdata.csv")
}

# DP1.3 loading the data & reading the file

storm <- read.csv("stormdata.csv.bz2",header=TRUE,sep=",")
summary(storm)
```

```{r}
names(storm)
```

## RESULTS
QUESTION 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Q1.1 Variable selection (reducing the data set to only needed columns and variables)
```{r}
variables<-c("EVTYPE","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP","CROPDMG","CROPDMGEXP")
strmdata<-storm[variables]

dim(strmdata)
```
```{r}
names(strmdata)
```
```{r}
Fatalities <- aggregate(FATALITIES ~ EVTYPE, data = strmdata, FUN = sum)
Top10_Fatalities <- Fatalities[order(-Fatalities$FATALITIES), ][1:10, ] 
Top10_Fatalities 
```
```{r}
Injuries <- aggregate(INJURIES ~ EVTYPE, data = strmdata, FUN = sum)
Top10_Injuries <- Injuries[order(-Injuries$INJURIES), ][1:10, ] 
Top10_Injuries 
```
```{r}
par(mfrow=c(1,2),mar=c(10,3,3,2))
barplot(Top10_Fatalities$FATALITIES,names.arg=Top10_Fatalities$EVTYPE,las=2,col="purple",ylab="fatalities",main="Top 10 fatalities")
barplot(Top10_Injuries$INJURIES,names.arg=Top10_Injuries$EVTYPE,las=2,col="purple",ylab="injuries",main="Top 10 Injuries")
```
Figure 1: The weather event responsbile for the highest fatalities and injuries is the ‘Tornado’
```{r}
unique(strmdata$PROPDMGEXP)
```
```{r}
strmdata$PROPEXP[strmdata$PROPDMGEXP == "K"] <- 1000
strmdata$PROPEXP[strmdata$PROPDMGEXP == "M"] <- 1e+06
strmdata$PROPEXP[strmdata$PROPDMGEXP == ""] <- 1
strmdata$PROPEXP[strmdata$PROPDMGEXP == "B"] <- 1e+09
strmdata$PROPEXP[strmdata$PROPDMGEXP == "m"] <- 1e+06
strmdata$PROPEXP[strmdata$PROPDMGEXP == "0"] <- 1
strmdata$PROPEXP[strmdata$PROPDMGEXP == "5"] <- 1e+05
strmdata$PROPEXP[strmdata$PROPDMGEXP == "6"] <- 1e+06
strmdata$PROPEXP[strmdata$PROPDMGEXP == "4"] <- 10000
strmdata$PROPEXP[strmdata$PROPDMGEXP == "2"] <- 100
strmdata$PROPEXP[strmdata$PROPDMGEXP == "3"] <- 1000
strmdata$PROPEXP[strmdata$PROPDMGEXP == "h"] <- 100
strmdata$PROPEXP[strmdata$PROPDMGEXP == "7"] <- 1e+07
strmdata$PROPEXP[strmdata$PROPDMGEXP == "H"] <- 100
strmdata$PROPEXP[strmdata$PROPDMGEXP == "1"] <- 10
strmdata$PROPEXP[strmdata$PROPDMGEXP == "8"] <- 1e+08

# Assigning '0' to invalid exponent strmdata
strmdata$PROPEXP[strmdata$PROPDMGEXP == "+"] <- 0
strmdata$PROPEXP[strmdata$PROPDMGEXP == "-"] <- 0
strmdata$PROPEXP[strmdata$PROPDMGEXP == "?"] <- 0

# Calculating the property damage value
strmdata$PROPDMGVAL <- strmdata$PROPDMG * strmdata$PROPEXP


# Q2.3 Defining & Calcuating [ Crop Damage ]

## Crop damage exponents for each level listed out & assigned those values for the crop exponent data. 
## Invalid data was excluded by assigning the value as '0'. 
## Then, the crop damage value was calculated by multiplying the crop damage and crop exponent value.

unique(strmdata$CROPDMGEXP)
```
```{r}
# Assigning values for the crop exponent strmdata 
strmdata$CROPEXP[strmdata$CROPDMGEXP == "M"] <- 1e+06
strmdata$CROPEXP[strmdata$CROPDMGEXP == "K"] <- 1000
strmdata$CROPEXP[strmdata$CROPDMGEXP == "m"] <- 1e+06
strmdata$CROPEXP[strmdata$CROPDMGEXP == "B"] <- 1e+09
strmdata$CROPEXP[strmdata$CROPDMGEXP == "0"] <- 1
strmdata$CROPEXP[strmdata$CROPDMGEXP == "k"] <- 1000
strmdata$CROPEXP[strmdata$CROPDMGEXP == "2"] <- 100
strmdata$CROPEXP[strmdata$CROPDMGEXP == ""] <- 1

# Assigning '0' to invalid exponent strmdata
strmdata$CROPEXP[strmdata$CROPDMGEXP == "?"] <- 0

# calculating the crop damage 
strmdata$CROPDMGVAL <- strmdata$CROPDMG * strmdata$CROPEXP


# Q2.4 Property Damage Summary

## Procedure = aggregate the property damage by the event type and sort the output it in descending order

prop <- aggregate(PROPDMGVAL~EVTYPE,data=strmdata,FUN=sum,na.rm=TRUE)
prop <- prop[with(prop,order(-PROPDMGVAL)),]
prop <- head(prop,10)
print(prop)
```
```{r}
# Q2.5 Crop Damage Summary

## Procedure = aggregate the crop damage by the event type and sort the output it in descending order

crop <- aggregate(CROPDMGVAL~EVTYPE,data=strmdata,FUN=sum,na.rm=TRUE)
crop <- crop[with(crop,order(-CROPDMGVAL)),]
crop <- head(crop,10)
print(crop)
```
```{r}
# Q2.6 Plot of Top 10 Property & Crop damages by Weather Event Types ( Economic Consequences )

##plot the graph showing the top 10 property and crop damages

par(mfrow=c(1,2),mar=c(11,3,3,2))
barplot(prop$PROPDMGVAL/(10^9),names.arg=prop$EVTYPE,las=2,col="gold",ylab="Prop.damage(billions)",main="Top10 Prop.Damages")
barplot(crop$CROPDMGVAL/(10^9),names.arg=crop$EVTYPE,las=2,col="gold",ylab="Crop damage(billions)",main="Top10 Crop.Damages")
```
Figure 2: ‘Floods’ are responsbile for the highest property damage while ‘droughts’ cause the greatest crop damage.

### **SUMMARY**

  *  Tornados are responsible for the maximum number of fatalities and injuries, followed by Excessive Heat for fatalities and Thunderstorm wind for injuries.

  *  Floods are responsbile for maximum property damage, while Droughts cause maximum crop damage. Second major events that caused the maximum damage was Hurricanes/Typhoos for property damage and Floods for crop damage.

